---
title: "heart-rate-visualization"
author: "Adam Resnick"
date: "5/21/2020"
output:
  html_document: default
  pdf_document: default
---

## Project Overview

For this project, I wanted to look at workout data collected from my Fitbit. I investigated the relationship between HR and speed, as created some cool visualizations.

### Reading in data

```{r read}
wo1RAW <-read.csv("2018-08-07 - 0629.csv", header=TRUE, stringsAsFactors = FALSE)
wo2RAW <-read.csv("2018-08-09 - 1958.csv", header=TRUE, stringsAsFactors = FALSE)
wo3RAW <-read.csv("2018-08-13 - 1705.csv", header=TRUE, stringsAsFactors = FALSE)
```

## Loading Libraries

I used the tidyverse library, which includes ggplot, dplyr, and more top R libraries.

```{r libraries}
library(tidyverse)
```

### Data Formatting, Data Cleaning, Speed Calculation

After reading in the datasets, I got rid of the columns I wasn't interested in, leaving the GPS data, heartrate, and distance. I then formatted the time and elliminated any rows with missing data.

```{r dataformatting}
wo1 <- wo1RAW

#deleting unnecessary columns in wo1
wo1$Sport <- NULL; wo1$ï..Sport <- NULL; wo1$ns1.TriggerMethod <- NULL; wo1$ns1.Name <- NULL; wo1$ns1.TotalTimeSeconds <- NULL
wo1$ns1. <- NULL; wo1$ns1.DistanceMeters <- NULL; wo1$ns1.UnitId <- NULL; wo1$ns1.ProductID <- NULL
wo1$ns1.Calories <- NULL; wo1$ns1.Id <- NULL; wo1$StartTime <- NULL; wo1$ns1.Intensity <- NULL

#renaming columns in wo1
names(wo1)[1] <- "DateTime"; names(wo1)[2] <- "Latitude"; names(wo1)[3] <- "Longitude"; names(wo1)[4] <- "Altitude"
names(wo1)[5] <- "Distance"; names(wo1)[6] <- "HR"

#Formatting time
wo1$DateTime <- sub(pattern="T", replace=" ", wo1$DateTime)
wo1$DateTime <- sub(pattern=".00.*", replace="", wo1$DateTime)
wo1$DateTime <- as.POSIXct(wo1$DateTime, format = "%Y-%m-%d %H:%M:%S")
wo1$Date <- sub(pattern=" .*", replace="", wo1$DateTime)
wo1$Date <- as.Date(wo1$Date)
wo1$Time <- sub(pattern=".* ", replace="", wo1$DateTime)

#Eliminating rows with any missing data
wo1 <- wo1[complete.cases(wo1),]
```

### Speed Calculation and Data Cleaning

I calculated speed using the distance and time variables. I then got rid of any junk data, such as negative speeds and speeds inconsistent with the workout I was doing (AKA: speeds over 15mph).

```{r speedcalc}
#Adding Speed
wo1$Speed <- rep(NA, length(wo1$DateTime))
for(i in 2:length(wo1$DateTime)){
  wo1$Speed[i] <- (wo1$Distance[i]-wo1$Distance[i-1])*0.0006213712/(as.numeric((wo1$DateTime[i]-wo1$DateTime[i-1]))/3600)
}

#Cleaning up dataframe
wo1 <- wo1[wo1$Speed>0,]        #Gets rid of rows with negative speed (not possible)
wo1 <- wo1[wo1$Speed<15,]        #Gets rid of rows with unrealistic speed
wo1 <- wo1[2:length(wo1$DateTime),]
```

### Saving Data as .csv file

While it isn't necessary to save the data as a .csv, I chose to do so just so I had a snapshot of where I was in the project.

```{r savecsv}
write.csv(wo1, file = ("Workout1.csv"), row.names = FALSE)
```


### Plotting data

I chose to create several plots and save them as .png files, which you can observe below. 

```{r plots}
#Plotting Lat, Long, HR for Workout 1
ggplot(data=wo1, aes(x=Latitude, y=Longitude, size=HR))+geom_point()+ggtitle("Workout 1 - Mapped HR")
ggsave("Workout1_HR.png", width = 6, height = 6, units = "in",
       dpi = 300)
#Plotting Lat, Long, Speed for Workout 1
ggplot(data=wo1, aes(x=Latitude, y=Longitude, color=-Speed))+geom_point()+ggtitle("Workout 1 - Mapped Speed")
ggsave("Workout1_Speed.png", width = 6, height = 6, units = "in",
       dpi = 300)
#Plotting Lat, Long, Speed, and HR for Workout 1
ggplot(data=wo1, aes(x=Latitude, y=Longitude, color=-Speed, size=HR))+geom_point()+ggtitle("Workout 1 - Mapped HR vs Speed")
ggsave("Workout1_HRSpeed.png", width = 6, height = 6, units = "in",
       dpi = 300)
#Plotting HR vs Speed for Workout 1
ggplot(data=wo1, aes(x=Speed, y=HR))+geom_point()+geom_smooth(method="lm", color="blue")+xlab("Speed [mph]")+ylab("HR [BPM]")+ggtitle("Workout 1 - HR vs Speed")
ggsave("Workout1_HRvsSpeed.png", width = 6, height = 6, units = "in",
       dpi = 300)

```

## Analyzing Workouts 2 and 3.

Workouts 2 and 3 were analyzed using the same techniques as Workout 1. For the sake of brevity, I will not include the code and only show the plots.


```{r pressure, echo=FALSE}
########################################################
# Workout 2 Code
########################################################

wo2 <- wo2RAW

#deleting unnecessary columns in wo2
wo2$Sport <- NULL; wo2$ï..Sport <- NULL; wo2$ns1.TriggerMethod <- NULL; wo2$ns1.Name <- NULL; wo2$ns1.TotalTimeSeconds <- NULL
wo2$ns1. <- NULL; wo2$ns1.DistanceMeters <- NULL; wo2$ns1.UnitId <- NULL; wo2$ns1.ProductID <- NULL
wo2$ns1.Calories <- NULL; wo2$ns1.Id <- NULL; wo2$StartTime <- NULL; wo2$ns1.Intensity <- NULL

#renaming columns in wo2
names(wo2)[1] <- "DateTime"; names(wo2)[2] <- "Latitude"; names(wo2)[3] <- "Longitude"; names(wo2)[4] <- "Altitude"
names(wo2)[5] <- "Distance"; names(wo2)[6] <- "HR"

#Formatting time
wo2$DateTime <- sub(pattern="T", replace=" ", wo2$DateTime)
wo2$DateTime <- sub(pattern=".00.*", replace="", wo2$DateTime)
wo2$DateTime <- as.POSIXct(wo2$DateTime, format = "%Y-%m-%d %H:%M:%S")
wo2$Date <- sub(pattern=" .*", replace="", wo2$DateTime)
wo2$Date <- as.Date(wo2$Date)
wo2$Time <- sub(pattern=".* ", replace="", wo2$DateTime)

#Eliminating rows with any missing data
wo2 <- wo2[complete.cases(wo2),]

#Adding Speed
wo2$Speed <- rep(NA, length(wo2$DateTime))
for(i in 2:length(wo2$DateTime)){
  wo2$Speed[i] <- (wo2$Distance[i]-wo2$Distance[i-1])*0.0006213712/(as.numeric((wo2$DateTime[i]-wo2$DateTime[i-1]))/3600)
}

#Cleaning up dataframe
wo2 <- wo2[wo2$Speed>0,]        #Gets rid of rows with negative speed (not possible)
wo2 <- wo2[wo2$Speed<15,]        #Gets rid of rows with unrealistic speed
wo2 <- wo2[2:length(wo2$DateTime),]

#Saving Workout2 as CSV
write.csv(wo2, file = ("Workout2.csv"), row.names = FALSE)

#Plotting Lat, Long, HR for Workout 2
ggplot(data=wo2, aes(x=Latitude, y=Longitude, size=HR))+geom_point()+ggtitle("Workout 2 - Mapped HR")
ggsave("Workout2_HR.png", width = 6, height = 6, units = "in",
       dpi = 300)
#Plotting Lat, Long, Speed for Workout 2
ggplot(data=wo2, aes(x=Latitude, y=Longitude, color=-Speed))+geom_point()+ggtitle("Workout 2 - Mapped Speed")
ggsave("Workout2_Speed.png", width = 6, height = 6, units = "in",
       dpi = 300)
#Plotting Lat, Long, Speed, and HR for Workout 2
ggplot(data=wo2, aes(x=Latitude, y=Longitude, color=-Speed, size=HR))+geom_point()+ggtitle("Workout 2 - Mapped HR vs Speed")
ggsave("Workout2_HRSpeed.png", width = 6, height = 6, units = "in",
       dpi = 300)


########################################################
# Workout 3 Code
########################################################

wo3 <- wo3RAW

#deleting unnecessary columns in wo3
wo3$Sport <- NULL; wo3$ï..Sport <- NULL; wo3$ns1.TriggerMethod <- NULL; wo3$ns1.Name <- NULL; wo3$ns1.TotalTimeSeconds <- NULL
wo3$ns1. <- NULL; wo3$ns1.DistanceMeters <- NULL; wo3$ns1.UnitId <- NULL; wo3$ns1.ProductID <- NULL
wo3$ns1.Calories <- NULL; wo3$ns1.Id <- NULL; wo3$StartTime <- NULL; wo3$ns1.Intensity <- NULL

#renaming columns in wo3
names(wo3)[1] <- "DateTime"; names(wo3)[2] <- "Latitude"; names(wo3)[3] <- "Longitude"; names(wo3)[4] <- "Altitude"
names(wo3)[5] <- "Distance"; names(wo3)[6] <- "HR"

#Formatting time
wo3$DateTime <- sub(pattern="T", replace=" ", wo3$DateTime)
wo3$DateTime <- sub(pattern=".00.*", replace="", wo3$DateTime)
wo3$DateTime <- as.POSIXct(wo3$DateTime, format = "%Y-%m-%d %H:%M:%S")
wo3$Date <- sub(pattern=" .*", replace="", wo3$DateTime)
wo3$Date <- as.Date(wo3$Date)
wo3$Time <- sub(pattern=".* ", replace="", wo3$DateTime)

#Eliminating rows with any missing data
wo3 <- wo3[complete.cases(wo3),]

#Adding Speed
wo3$Speed <- rep(NA, length(wo3$DateTime))
for(i in 2:length(wo3$DateTime)){
  wo3$Speed[i] <- (wo3$Distance[i]-wo3$Distance[i-1])*0.0006213712/(as.numeric((wo3$DateTime[i]-wo3$DateTime[i-1]))/3600)
}

#Cleaning up dataframe
wo3 <- wo3[wo3$Speed>0,]        #Gets rid of rows with negative speed (not possible)
wo3 <- wo3[wo3$Speed<15,]        #Gets rid of rows with unrealistic speed
wo3 <- wo3[2:length(wo3$DateTime),]

#Saving Workout3 as CSV
write.csv(wo3, file = ("Workout3.csv"), row.names = FALSE)

#Plotting Lat, Long, HR for Workout 3
ggplot(data=wo3, aes(x=Latitude, y=Longitude, size=HR))+geom_point()+ggtitle("Workout 3 - Mapped HR")
ggsave("Workout3_HR.png", width = 6, height = 6, units = "in",
       dpi = 300)
#Plotting Lat, Long, Speed for Workout 3
ggplot(data=wo3, aes(x=Latitude, y=Longitude, color=-Speed))+geom_point()+ggtitle("Workout 3 - Mapped Speed")
ggsave("Workout3_Speed.png", width = 6, height = 6, units = "in",
       dpi = 300)
#Plotting Lat, Long, Speed, and HR for Workout 3
ggplot(data=wo3, aes(x=Latitude, y=Longitude, color=-Speed, size=HR))+geom_point()+ggtitle("Workout 3 - Mapped HR vs Speed")
ggsave("Workout3_HRSpeed.png", width = 6, height = 6, units = "in",
       dpi = 300)
```

### Regression Analysis

Finally, I performed a regression analysis on each of the workout's HR vs Speed. I also compared log(HR) vs Speed and HR vs log(Speed), but straight HR vs Speed produced the highest R-squared values for each of the workouts.

#### Workout 1

```{r regression1}
ggplot(data=wo1, aes(x=Speed, y=HR))+geom_point()+geom_smooth(method="lm", color="blue")+xlab("Speed [mph]")+ylab("HR [BPM]")+ggtitle("Workout 1 - HR vs Speed")
mod.1 <- lm(formula=HR~Speed, data=wo1)
summary(mod.1)
```

#### Workout 2

You'll notice that Workout 2 has a very low R-squared value. This is a result of very little variance of speed (I walked all of it) and thus little variance in HR. While the overall trend mirrors that of the other two workouts, the lack of any real variance in speed makes for a poor training set. It's included here purely for the sake of discussion.

```{r regression2}
ggplot(data=wo2, aes(x=Speed, y=HR))+geom_point()+geom_smooth(method="lm", color="blue")+xlab("Speed [mph]")+ylab("HR [BPM]")+ggtitle("Workout 2 - HR vs Speed")
mod.2 <- lm(formula=HR~Speed, data=wo2)
summary(mod.2)
```

#### Workout 3

```{r regression3}
ggplot(data=wo3, aes(x=Speed, y=HR))+geom_point()+geom_smooth(method="lm", color="blue")+xlab("Speed [mph]")+ylab("HR [BPM]")+ggtitle("Workout 3 - HR vs Speed")
mod.3 <- lm(formula=HR~Speed, data=wo3)
summary(mod.3)
```

### Summary

This was a fun project to work on that reinforced the basic R concepts from this class. Now that I have the data cleaning operations set up, I can upload future workout data quite easily.

### Further Work

I would investigate ways to replace any missing data. While this is not necessary for the analysis, it would improve the readibility of the plots that include latitude and longitude.

Additionally, I plan on recording data for more speeds, not just ~4mph and ~7.5mph. Ideally, I would start out at 4mph and slowly increase my speed to ~8mph. This would give a better understanding of how HR truly differs with speed. I'd be curious to see if the relationship would be linear or if it would be better fit by something polynomial.

